Experience Report: Haskell in Computational Biology

### 2012-06-21 mrfy_experience_report.tex
A few crude changes, and many breadcrumbs (in remarks) that need to be swept up and dealt with.

A~hidden Markov model carries probabilities on some states and on all
state transitions.
Both the probabilities and the states are determined by the alignment:
\begin{itemize}
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{match state}~$M_j$.
The match state contains a table $e_{M_j}(x)$ which gives the
 probability that a homologous protein has amino acid~$x$ in
 column~$j$.
This probability is called an \emph{emission} probability.
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{deletion state}~$D_j$.
The deletion state determines the probability that the query sequence
has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
\item 
For each column~$j$ of the alignment, the hidden Markov model has an
\emph{insertion state}~$I_j$.
The insertion state contains a table $e_{I_j}(x)$ which gives the
probability that a homologous protein has gained amino acid~$x$ by
insertion at column~$j$.
\end{itemize}
The hidden Markov model also has distinguished \emph{begin} and \emph{end} states.
We~represent a state using an algebraic data type:
\begin{smallverbatim}
data HMMState = Mat | Ins | Del | Beg | End
\end{smallverbatim}

@@

Given the \texttt{Scored} abstraction, we can easily implement
Viterbi's equations for a match state.
To~implement the equation at the top right of \figref{viterbi},
we~choose the best solution
from among the three predecessor states;
we~use~\texttt{/+/} to add transition and emission scores,
and we use \texttt{fmap} to add \texttt{Mat} to the list of states
encountered up to this point:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\smallverbatiminput{vfix}
Functions \texttt{tProb} and \texttt{edge} combine to select the
appropriate transition score (one of the $a'$~values).
Function~\verb+vee''+ is the \emph{memoized} version of~\verb+vee'+,
so
the call to~\verb+vee''+ produces the same result as a recursive call
to~\verb+vee'+, but faster:
\smallverbatiminput{memo}
Functions \texttt{Memo.memo3} and \texttt{Memo.arrayRange} come from
Luke Palmer's
\path{Data.MemoCombinators} package.
The value
\texttt{numNodes} represents~$\alignwidth$,
and \texttt{seqlen} represents~$N$.

@@

Our implementations of hidden Markov models and Viterbi's algorithm
duplicate existing functionality.
In~MRFy, they are used to implement new functionality.

@@

We~pass the
search function a \texttt{Scorer}, which runs Viterbi's algorithm.
\smallverbatiminput{orig-scoredecl}
A~search heuristic is characterized by four functions, which are
bundled into a record called \texttt{SearchStrategy}.
\smallfuzzverbatiminput{2.3pt}{orig-strategy}
In this record,
\begin{itemize}
\item
Function \texttt{pt0} produces the first generation of placements.
\item
Function \texttt{nextGen} produces a new generation of placements from
an existing generation.
Because scoring a placement is expensive, 
and some placements may persist from the old generation to the new
generation, \texttt{nextGen} works with \emph{scored} placements, not
bare placements.
\item
A~new generation produced by \texttt{nextGen} may be no better
than the generation it replaces.
When it is first produced, the new generation is ``on trial.''
The \texttt{accept} function decides whether to keep a new generation
or to try \texttt{nextGen} again with a different seed.
The \texttt{accept} function considers a list of type
\texttt{[Scored~Age]}.
At~the head of this list is the score of the best
placement of the trial generation, together with the \texttt{Age} of the trial
generation, which is the total number of generations tried.
The remaining elements of the list hold the history of scores and ages for
previously \emph{accepted} generations.
(The list omits generations that didn't survive the \texttt{accept} trial.)
For convenience, the \texttt{accept} function also receives
the \texttt{Age} of the trial generation.
\item
Finally, the \texttt{quit} function decides when to stop the search.
\ifnotcutting
It~too considers the list of accepted generations and the age of the current
generation. 
\fi
\end{itemize}
\seclabel{search}%
We hope you find the search simple and straightforward:
\smallfuzzverbatiminput{11pt}{orig-search}

@@

MRFy calls \verb+vee'+ several
 times
for each placement in each search generation.
Calls on query sequences between beta strands are independent,
as are calls on different placements.
We~found them easy to parallelize
using
\texttt{Control.Parallel}:
we substituted
\texttt{parmap rseq} for \texttt{map}.
Even when there is just one placement per generation,
MRFy can keep six cores busy.
Given multiple placements per generation,
it~can use forty-eight cores.

@@

We~think \verb+vee'+ is beautiful, but we wrote some ugly code, too,
because we lack skills and knowledge in two areas.
The first is profiling.
GHC's profiling is based on cost centers \cite{sansom-pj},
and although GHC 
will happily add cost centers to top-level 
functions, much of our code is in nested functions.
We~know of no better technique than to add cost-center annotations to
each guard clause and nested function,
but these annotations made our code so ugly that we felt compelled to
remove them as soon as possible.
And without the annotations, GHC's call-site and allocation
profiles did not give us enough information to be useful.
We~miss valgrind, whose callgrind profiler gives useful information
without \emph{any} annotations in the source.

@@

When performance has mattered, members of our group, like other
computational biologists, have used~C++.
We~compare our experience with three tools:
\begin{itemize}
\item
Matt~\cite{Menke:2008wu} is used to create alignments like that shown
in \figref{alignment}.
It~comprises about 12,000 lines of~C++.
The only external tools or libraries it uses are \texttt{zlib}
and \texttt{OpenMP}, and its initial development took two years.
\item
SMURF
\cite{Menke:2010ti} is used to detect homologous proteins in the presence
of paired beta strands.
It~comprises about 9,000 lines of~C++, of which about 1,000~lines are
shared with~Matt.
It~uses no external tools or libraries, and
its initial development took a year and a half.
\item
MRFy is used to detect homologous proteins in the presence of paired
beta strands.
It~comprises about 1,200 lines of~Haskell.
It~uses several external tools and libraries, of which the most
notable are PADS-Haskell, the BioHaskell bioinformatics library, and
the libraries \path{Data.MemoCombinators}, \path{Control.Parallel},
and \path{Data.Vector}.
Its initial development took about three months.
%%%% \remark{Others were mostly full-time work, but no need to distinguish.}
\end{itemize}
Like much research software, all three of these tools were written
 in~haste.
We~have experience modifying the older tools.

@@

%
% An Experience Report is about what you actually did, not what you
% wish to do :-)
%
%%  There are two major research ideas we wish to add to MRFy eventually, one of which
%%  \remark{how much should I say about profiles and alpha helices? ---NMD}
%%  involves looking at another type of nonlocal dependency (\textit{alpha
%%  helices} in contrast 
%%  to the current beta strands).
%%  We expect that thanks to algebraic data types and typeclasses, these
%%  enhancements will be at 
%%  most a few weeks of work; we do not believe these enhancements would
%%  even be feasible in SMURF. 
%%  Indeed, Daniels tried to add another of these enhancements to SMURF
%%  two years ago and gave up due 
%%  to the difficulty of the implementation.
%%  \remark{The above needs a lot of revision. It is a shitty first
%%  draft of a /P/. ---NMD}

@@

We~can also make a preliminary comparison of performance.
SMURF can parse a simple protein structure, optimally, in under 10~seconds.
It~can parse some more complex structures in 2--3~minutes.
But when asked to parse the most complex structures, it runs out of
memory and has to be killed, even on a system with 128GB of~RAM.
By~contrast, MRFy can parse any of these structures, approximately,
in 30~seconds.
This performance
meets our goals, and it means that MRFy can be used to scan whole genomes.

>>>>

A~hidden Markov model carries probabilities on some states and on all
state transitions.
Both the probabilities and the states are determined by the alignment:
\begin{itemize}
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{match state}~$M_j$.
The match state contains a table $e_{M_j}(x)$ which gives the
 probability that a homologous protein has amino acid~$x$ in
 column~$j$.
This probability is called an \emph{emission} probability.
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{deletion state}~$D_j$.
The deletion state determines the probability that the query sequence
has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
\item 
For each column~$j$ of the alignment, the hidden Markov model has an
\emph{insertion state}~$I_j$.
The insertion state contains a table $e_{I_j}(x)$ which gives the
probability that a homologous protein has gained amino acid~$x$ by
insertion at column~$j$.
\end{itemize}
The hidden Markov model also has distinguished \emph{begin} and \emph{end} states.
We~represent a state using an algebraic data type:
\remark{we promised a better name for HMMState. Thoughts?}
\begin{smallverbatim}
data HMMState = Mat | Ins | Del | Beg | End
\end{smallverbatim}

@@

Given the \texttt{Scored} abstraction, we can easily implement
Viterbi's equations for a match state.
To~implement the equation at the top right of \figref{viterbi},
we~choose the best solution
from among the three predecessor states;
we~use~\texttt{/+/} to add transition and emission scores,
and we use \texttt{fmap} to add \texttt{Mat} to the list of states
encountered up to this point:\remark{we promised to characterize the missing cases}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\smallverbatiminput{vfix}
Functions \texttt{tProb} and \texttt{edge} combine to select the
appropriate transition score (one of the $a'$~values).
Function~\verb+vee''+ is the \emph{memoized} version of~\verb+vee'+,
so
the call to~\verb+vee''+ produces the same result as a recursive call
to~\verb+vee'+, but faster:
\smallverbatiminput{memo}
Functions \texttt{Memo.memo3} and \texttt{Memo.arrayRange} come from
Luke Palmer's
\path{Data.MemoCombinators} package.
The value
\texttt{numNodes} represents~$\alignwidth$,
and \texttt{seqlen} represents~$N$.

@@

Our implementations of hidden Markov models and Viterbi's algorithm
duplicate existing functionality; these same algorithms were
implemented in C++ for HMMER and SMURF.
This reimplementation is a cost of using Haskell, and one that
other researchers may have to pay as well.
However, the authors of SMURF and HMMER implemented these algorithms
anew, even though existing implementations existed in C++.
In~MRFy, we use these algorithms as building blocks to implement
new functionality.

@@

We~pass the
search function a \texttt{Scorer}, which runs Viterbi's algorithm.
\smallverbatiminput{orig-scoredecl}
A~search heuristic is characterized by four functions, which are
bundled into a record called \texttt{SearchStrategy}.
\smallfuzzverbatiminput{2.3pt}{orig-strategy}
In this record,
\begin{itemize}
\item
Function \texttt{pt0} produces the first generation of placements.
\item
Function \texttt{nextGen} produces a new generation of placements from
an existing generation.
Because scoring a placement is expensive, 
and some placements may persist from the old generation to the new
generation, \texttt{nextGen} works with \emph{scored} placements, not
bare placements.
\item
A~new generation produced by \texttt{nextGen} may be no better
than the generation it replaces.
When it is first produced, the new generation is ``on trial.''
The \texttt{accept} function decides whether to keep a new generation
or to try \texttt{nextGen} again with a different seed.
The \texttt{accept} function considers a list of type
\texttt{[Scored~Age]}.
At~the head of this list is the score of the best
placement of the trial generation, together with the \texttt{Age} of the trial
generation, which is the total number of generations tried.
The remaining elements of the list hold the history of scores and ages for
previously \emph{accepted} generations.
(The list omits generations that didn't survive the \texttt{accept} trial.)
For convenience, the \texttt{accept} function also receives
the \texttt{Age} of the trial generation.
\item
Finally, the \texttt{quit} function decides when to stop the search.
\ifnotcutting
It~too considers the list of accepted generations and the age of the current
generation. 
\fi
\end{itemize}
This \texttt{SearchStrategy} interface somewhat resembles an object-oriented
style of programming.
However, it is \textit{easier} to create a SearchStrategy
in Haskell than using conventional object-oriented techniques, because
all the functions in a SearchStrategy are partial applications of
other functions.
If we were using conventional object-oriented
techniques, we would have to declare subclasses with instance
variables and would have to write code to initialize objects of those
subclasses.

\seclabel{search}%
We hope you find the search simple and straightforward:

\smallfuzzverbatiminput{11pt}{orig-search}

@@

MRFy calls \verb+vee'+ several
 times
for each placement in each search generation.
Calls on query sequences between beta strands are independent,
as are calls on different placements.
We~found them easy to parallelize
using
\texttt{Control.Parallel}:
we substituted
\texttt{parmap rseq} for \texttt{map}.
Even when there is just one placement per generation,
MRFy can keep six cores busy.
Given multiple placements per generation,
it~can use forty-eight cores.
However, we have found that MRFy performs best, at least on our
hardware, with between 12 and 16 cores. We have access to only
one 48-core system, and while MRFy will keep all 48 cores busy, it
performs better when given no more than 16 cores.\remark{give specific
numbers here}

MRFy's predecessor, SMURF, does not perform stochastic search,
and thus its overall run-time performance cannot be compared directly to MRFy.
However, both programs implement the Viterbi algorithm.
On an exceedingly large, contrived example, comprising 56,000 amino acids
(larger than any known protein) and 343 nodes, and ignoring all beta-strand
information, SMURF's Viterbi implementation completed in 0.497 seconds on a
2.6GHz Core i7.
On the same benchmark, MRFy's Viterbi implementation required 1 minute 57
seconds.
Despite the inferior runtime performance, MRFy still meets our performance
goals.
This is because the stochastic search algorithm relies on the Viterbi algorithm
applied to many, significantly smaller hidden Markov models than SMURF's
multidimensional programming must tackle.
MRFy computes an alignment for a 12-stranded ``beta sandwich'' structure in
generally less than one minute, while SMURF consumes over 16GB of memory
and does not terminate even after eight hours.
Had we implemented the SMURF algorithm in Haskell, we believe that performance
would be unacceptable.
However, for the algorithm underlying MRFy, we are satisfied with the runtime
performance of Haskell and GHC.

@@

We~think \verb+vee'+ is beautiful, but we wrote some ugly code, too,
because we lack skills and knowledge in two areas.
The first is profiling.
GHC's profiling is based on cost centers \cite{sansom-pj},
and for most of MRFy's development, we used GHC 7.0, which
will happily add cost centers to top-level 
functions only, while much of our code is in nested functions.
We~know of no better technique than to add cost-center annotations to
each guard clause and nested function,
but these annotations made our code so ugly that we felt compelled to
remove them as soon as possible.
And without the annotations, GHC's call-site and allocation
profiles did not give us enough information to be useful.
We~miss valgrind, whose callgrind profiler gives useful information
without \emph{any} annotations in the source.
Late in MRFy's development, we began using GHC 7.4, which provides much
better profiling tools.
We are thrilled by these tool developments.

@@

When performance has mattered, members of our group, like other
computational biologists, have used~C++.
We~compare our experience with three tools:
\begin{itemize}
\item
Matt~\cite{Menke:2008wu} is used to create alignments like that shown
in \figref{alignment}.
It~comprises about 12,000 lines of~C++.\remark{reviewers seemed to want a
table of this information, but that seems like a huge waste of space.}
The only external tools or libraries it uses are \texttt{zlib}
and \texttt{OpenMP}, and its initial development took two years.
\item
SMURF
\cite{Menke:2010ti} is used to detect homologous proteins in the presence
of paired beta strands.
It~comprises about 9,000 lines of~C++, of which about 1,000~lines are
shared with~Matt.
It~uses no external tools or libraries, and
its initial development took a year and a half.
\item
MRFy is used to detect homologous proteins in the presence of paired
beta strands.
It~comprises about 1,200 lines of~Haskell.
It~uses several external tools and libraries, of which the most
notable are Parsec, the BioHaskell bioinformatics library, and
the libraries \path{Data.MemoCombinators}, \path{Control.Parallel},
and \path{Data.Vector}.
Its initial development took about three months.
%%%% \remark{Others were mostly full-time work, but no need to distinguish.}
\end{itemize}
Like much research software, all three of these tools were written
 in~haste.
We~have experience modifying the older tools.

@@

@@

We~can also make a preliminary comparison of performance.
SMURF can parse a simple protein structure, optimally, in under 10~seconds.
It~can parse some more complex structures in 2--3~minutes.
But when asked to parse the most complex structures, it runs out of
memory and has to be killed, even on a system with 128GB of~RAM.
By~contrast, MRFy can parse any of these structures, approximately,
in 30~seconds.\remark{reconcile this vague paragraph with performance section above}
This performance
meets our goals, and it means that MRFy can be used to scan whole genomes.
